Reference Tracking Strategies of Native and Late Signers in Turkish Sign Language

### 2022-08-06 revised_manuscript.tex
updated files

Sign Languages are natural languages of the Deaf communities all around
the world. They are independent languages, which do not derive from
majority spoken language(s) but emerge naturally when deaf people form a
community. Turkish Sign Language (Türk İşaret Dili -- TİD) is and has
been the officially recognized language of the Deaf community in Turkey
since 2005 (see Nuhbalaoglu, 2018 for a brief review of the historical
development of the language). Deaf or hearing children acquiring a SL
within a family where there is one or more deaf caregivers display the
same developmental milestones as hearing children acquiring a spoken
language in a hearing family (Chamberlain et al., 1999; Meier and
Newport, 1990). However, reports indicate that over 90\% of all deaf
individuals are born into hearing families who do not know how to sign
(i.e., those with language deprivation) (Woll, 2013). This means that
the auditory modality used in the household is not accessible to many
deaf children. That is why the age of acquisition of a first functional
language varies greatly among deaf children, with many of whom being
deprived of their first languages typically until after age 7 when they
start formal education in a deaf school. First language deprivation or
delay has been found to have negative effects on certain cognitive and
linguistic measures among late signing children. To name a few, delayed
exposure to a first functional language is reported to negatively affect
deaf children and adults' morphosyntactic production and processing
(Boudreault and Mayberry, 2006; Cheng and Mayberry, 2021; Cormier et
al., 2012; Karadöller et al., 2017; Kayabaşı et al., 2020; Newport,
1990) and narrative abilities (e.g., Becker, 2009). In our overall
research, we are concerned with the effects of this late exposure in the
ultimate adult forms of TİD users.

@@

However, this particular observation is not peculiar to spoken
languages. There is an emerging body of literature that reports the
distribution of REs in signed narratives. Building on Gundel et al.
(1993)'s work, Swabey (2002) demonstrated that, like other spoken
languages such as Russian, Chinese, and Japanese, American Sign Language
(ASL) signers use bare nouns for introduced and re-introduced discourse
contexts whereas null marking is used for referents which are
``in-focus'' or very accessible to the addressee. Overt REs like
pronouns or nouns are disprefered for such referents. Swabey (2002)
concluded that ``native'' signers in her study abided by the Gricean
maxim of quantity and were thus economic in their reference tracking.
Similar findings have been reported for German Sign Language (Deutsche
Gebärdensprache - DGS) (Perniss and Özyürek, 2015) and British Sign
Language (BSL) (Morgan, 2005). It is important to note, however, that
these research either collapsed native and late signers into a single
group and labeled them as ``native-like'' or explicitly compared native
deaf signers with late deaf signers (with language deprivation) in terms
of narrative competence.

\hypertarget{signed-vs.-spoken-res}{%
\subsubsection{Signed vs.~Spoken REs}\label{signed-vs.-spoken-res}}

Commonly in both spoken and sign languages nouns and noun phrases are
used as REs. Spoken languages use pronouns, determines and
demonstratives where sign languages use a pointing sign, usually glossed
as IX, which assumes similar functions (Lillo-Martin and Klima, 1990).
In addition, the pointing sign is used to referentially anchor a
referent to a location in signing space. For instance, when the signer
introduces \textsc{elvan} for the first time, s/he can assign her to the
right. This referential locus can later be used by agreement verbs and
classifier predicates (more below). For instance, in (\ref{ex:2a}) there
are two referents, \textsc{elvan}, who is referentially-assigned to the
right side of the signer, represented with the ``A'' subscript and
\textsc{demet}, who is referentially-assigned to the left side of the
signer, represented with the ``B'' subscript. When the signer makes a
contrast in (\ref{ex:2b}), as a pronominal use, she points to the right
side to refer to ELVAN and to the left to refer to \textsc{demet}. We
schematically represent the referential locus assigned to the referents
in (\ref{ex:3}).

@@

Grammatical gender as in Romance languages or noun classes as in Bantu
languages are some reference tracking tools that are familiar from
spoken languages. A somehow less known reference-tracking tool in spoken
languages is \emph{verbal classifiers}, which morphologically attach to
the verb and often cross-reference a core argument in a clause
(Aikhenvald, 2017). Classifiers denote sets of objects based on semantic
criteria such as size, shape, body-part, texture and sometimes function.
Usually, the verbal classifier and the cross-referenced noun co-occur in
the same clause when the argument is introduced into discourse. The
classifier can track the argument on its own on subsequent occurrences.
In sign languages, a verbal classifier morpheme is encoded by a
\emph{handshape} that is articulated phonologically co-temporally with a
movement, which encodes the verbal meaning (Benedicto and Brentari,
2004; Supalla, 1990). These two infused morphemes form a complex
predicate. Some classifier handshapes and the set of objects that they
denote are presented below in (\ref{ex:4}).

@@

\begin{multicols}{2}
\begin{exe}
\ex
\begin{xlist}
\ex \label{ex:6a} Full noun \\\glll
\includegraphics[width=50pt]{pictures/1a_1.png}\\ \textsc{mouse\footnotemark} \\ \\ 
\ex \label{ex:6b} Noun modified by classifier \\\glll
\includegraphics[width=50pt]{pictures/1c_2.png} \includegraphics[width=50pt]{pictures/1c_1.png} \\ \textsc{soap} \textsc{soap.cl} \\ \\ 
\ex \label{ex:6c} Fingerspelled noun\\\glll 
\includegraphics[width=50pt]{pictures/1b_1.png} \includegraphics[width=50pt]{pictures/1b_2.png} \includegraphics[width=50pt]{pictures/1b_3.png}\\ \textsc{t-} \textsc{o-} \textsc{m} \\ \\
\ex \label{ex:6d} Noun modified by index (\textsc{IX})\\\glll 
\includegraphics[width=50pt]{pictures/1d_1.png} \includegraphics[width=50pt]{pictures/1d_2.png}\\ \textsc{ix} \textsc{cat} \\ \\
\end{xlist}
\end{exe}
\end{multicols}
\ea 
\ea \label{ex:7} Pronominal \\\glll
\includegraphics[width=50pt]{pictures/2_1.png} \includegraphics[width=50pt]{pictures/2_2.png} \\ \textsc{ix} \textsc{sick} \\ \\
\z \z
\begin{multicols}{4}
\begin{exe}
\ex \label{ex:8}
\begin{xlist}
\ex Whole entity \\\glll 
\includegraphics[width=50pt]{pictures/3e.png} \\ \textsc{fish.fall}\\ \\
\ex Body part \\\glll
\includegraphics[width=50pt]{pictures/3b.png} \\ \textsc{walk} \\ \\ 
\ex Handling \\\glll 
\includegraphics[width=50pt]{pictures/3c.png} \\ \textsc{hold} \\ \\
\ex Extension \\\glll 
\includegraphics[width=50pt]{pictures/3d_1.png} \includegraphics[width=50pt]{pictures/3a.png}\\ \textsc{plank} {\textsc{dh: mouse}} \\ {} {\textsc{ndh: tunnel}}\\ \\
\end{xlist}
\end{exe}
\end{multicols}
\begin{multicols}{3}
\begin{exe}
\ex
\begin{xlist}
\ex \label{ex:9a} Constructed action \\\glll 
\includegraphics[width=50pt]{pictures/4c.png} \\ \textsc{struggle} \\ \\

The IX sign (extended index finger) has various functions in sign
languages, one of which is the pronominal use which refers to
established nominals by pointing to a previously assigned abstract or
physical locus in the signing space when re-introducing a topic in
discourse (Emmorey, 1996). The same IX function has also been identified
for TİD (Sevinç, 2006). The example in (\ref{ex:7}) illustrates a
spatially-anchored pronominal reference for topic re-introduction. These
signs are thus rarely used for the first mention of an entity (Perniss
and Özyürek, 2015). Other interpretations consider pointing signs
without noun phrases as examples of demonstratives and not personal
pronouns (Koulidobrova and Lillo-Martin, 2016). It is important to
mention here that in some spoken languages (e.g., German and Dutch),
demonstratives can also refer to persons (Bosch and Umbach, 2007) and
behave similarly to personal pronouns. However, for the purposes of the
present study, we will refer to IX constructions as pronominal IX and
particularly examine their non-deictic or personal pronoun realizations
as a possible discourse strategy to mainly mark topic shift among native
and late signers of TİD.

Classifier predicates are highly complex morphological constructions
which are very common in sign languages (Supalla, 1986). These are
manually signed iconic representations of animate or inanimate objects,
and they embody several properties pertaining to the nominal referent
including its location in the signing space, movement, path, size, and
shape (Talmy, 2003). As in most sign languages hitherto studied,
classifier constructions are also commonly found in TİD (see Dikyuva et
al., 2015 for a review). Following the classification of
Engberg-Pedersen (1993), we distinguish between four types of classifier
constructions. Whole entity classifiers (WECL) represent the entirety of
any referent whereas body part classifiers (BPCL) only reflect a
specific part of an animate entity (e.g., limbs, head, tongue, teeth,
and so forth). Handling classifiers (HCL) occur with transitive verbs
and depict how an object is handled or manipulated by an agent. On this
ground, an HCL\footnote{As discussed before, an HCL is transitive in
  nature and depicts the handling action performed by ananimate agent.
  This is why most research tend to consider HCLs as an extension to
  constructed action (Cormier et al., 2013; Dudis, 2004).} depicts the
manner of the referring action while the former two constructions encode
the location of the entity as well as its motion in space. Extension
classifiers (ExtCL), on the other hand, denote the size and shape of an
entity by following a trajectory to trace its shape. We exemplify these
constructions in (\ref{ex:8}).

@@

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\vadjust pre{\hypertarget{ref-ahn2019}{}}%
Ahn, Dorothy, 2019. The determinacy scale: A competition mechanism for
anaphoric expressions (PhD thesis).

@@

@@

@@

>>>>

Sign Languages are natural languages of the Deaf communities all around
the world. They are independent languages, which do not derive from
majority spoken language(s) but emerge naturally when deaf people form a
community. Turkish Sign Language (Türk İşaret Dili -- TİD) is and has
been the officially recognized language of the Deaf community in Turkey
since 2005 (see Nuhbalaoglu, 2018 for a brief review of the historical
development of the language). Deaf or hearing children acquiring a SL
within a family where there is one or more deaf caregivers display the
same developmental milestones as hearing children acquiring a spoken
language in a hearing family (Chamberlain et al., 1999; Meier and
Newport, 1990). However, reports indicate that over 90\% of all deaf
individuals are born into hearing families who do not know how to sign
(i.e., those with language deprivation) (Woll, 2013). This means that
the auditory modality used in the household is not accessible to many
deaf children. That is why the age of acquisition of a first functional
language varies greatly among deaf children, with many of whom being
deprived of their first languages typically until after age 7 when they
start formal education in a deaf school. Despite this, many of the early
intervention programs for late signing children prioritize spoken
language development and discourage the use of a sign language
(Kemaloğlu and Kemaloğlu, 2012). Ministry of National Education (2020)
identifies 72 deaf schools in Turkey for the K-12 level. Even when deaf
children with hearing parents start their formal education in one of
these schools, they do not receive proper instruction in TİD since the
teachers themselves are mainly hearing and only have a very rudimentary
knowledge of a sign language (Ilkbasaran, 2015). First language
deprivation or delay has been found to have negative effects on certain
cognitive and linguistic measures among late signing children. To name a
few, delayed exposure to a first functional language is reported to
negatively affect deaf children and adults' morphosyntactic production
and processing (Boudreault and Mayberry, 2006; Cheng and Mayberry, 2021;
Cormier et al., 2012; Karadöller et al., 2017; Kayabaşı et al., 2020;
Newport, 1990) and narrative abilities (e.g., Becker, 2009). In our
overall research, we are concerned with the effects of this late
exposure in the ultimate adult forms of TİD users.

@@

However, this particular observation is not peculiar to spoken
languages. There is an emerging body of literature that reports the
distribution of REs in signed narratives (Ahlgren and Bergman, 1994;
Czubek, 2017; Frederiksen and Mayberry, 2016; Hodge et al., 2019;
Morgan, 2005; Perniss and Özyürek, 2015; Swabey, 2002). Building on
Gundel et al. (1993)'s work, Swabey (2002) demonstrated that, like other
spoken languages such as Russian, Chinese, and Japanese, American Sign
Language (ASL) signers use bare nouns for introduced and re-introduced
discourse contexts whereas null marking is used for referents which are
``in-focus'' or very accessible to the addressee. Overt REs like
pronouns or nouns are disprefered for such referents. Swabey (2002)
concluded that ``native'' signers in her study abided by the Gricean
maxim of quantity and were thus economic in their reference tracking.
Similar findings have been reported for German Sign Language (Deutsche
Gebärdensprache - DGS) (Perniss and Özyürek, 2015) and British Sign
Language (BSL) (Morgan, 2005). It is important to note, however, that
these research either collapsed native and late signers into a single
group and labeled them as ``native-like'' or explicitly compared native
deaf signers with late deaf signers (with language deprivation) in terms
of narrative competence.

The IX sign (extended index finger) has various functions in sign
languages, one of which is the pronominal use which refers to
established nominals by pointing to a previously assigned abstract or
physical locus in the signing space when re-introducing a topic in
discourse (Emmorey, 1996). The same IX function has also been identified
for TİD (Sevinç, 2006). The example in (\ref{ex:7}) illustrates a
spatially-anchored pronominal reference for topic re-introduction. These
signs are thus rarely used for the first mention of an entity (Perniss
and Özyürek, 2015). Other interpretations consider pointing signs
without noun phrases as examples of demonstratives and not personal
pronouns (Koulidobrova and Lillo-Martin, 2016). However, for the
purposes of the present study, we will refer to IX constructions as
pronominal IX and particularly examine their non-deictic or personal
pronoun realizations as a possible discourse strategy to mainly mark
topic shift among native and late signers of TİD.

Classifier predicates are highly complex morphological constructions
which are very common in sign languages (Supalla, 1986). These are
manually signed iconic representations of animate or inanimate objects,
and they embody several properties pertaining to the nominal referent
including its location in the signing space, movement, path, size, and
shape (Talmy, 2003). As in most sign languages hitherto studied,
classifier constructions are also commonly found in TİD (see Dikyuva et
al., 2015 for a review). Following the classification of
Engberg-Pedersen (1993), we distinguish between four types of classifier
constructions. Whole entity classifiers (WECL) represent the entirety of
any referent whereas body part classifiers (BPCL) only reflect a
specific part of an animate entity (e.g., limbs, head, tongue, teeth,
and so forth). Handling classifiers (HCL\footnote{As discussed before,
  an HCL is transitive in nature and depicts the handling action
  performed by ananimate agent. This is why most research tend to
  consider HCLs as an extension to constructed action (Cormier et al.,
  2013; Dudis, 2004).}) occur with transitive verbs and depict how an
object is handled or manipulated by an agent. Extension classifiers
(ExtCL), on the other hand, denote the size and shape of an entity by
following a trajectory to trace its shape. We exemplify these
constructions in (\ref{ex:8}).

\begin{multicols}{2}
\begin{exe}
\ex
\begin{xlist}
\ex \label{ex:6a} Full noun \\\glll
\includegraphics[width=50pt]{pictures/1a_1.png}\\ \textsc{mouse\footnotemark} \\ \\ 
\ex \label{ex:6b} Noun modified by classifier \\\glll
\includegraphics[width=50pt]{pictures/1c_2.png} \includegraphics[width=50pt]{pictures/1c_1.png} \\ \textsc{soap} \textsc{soap.cl} \\ \\ 
\ex \label{ex:6c} Fingerspelled noun\\\glll 
\includegraphics[width=50pt]{pictures/1b_1.png} \includegraphics[width=50pt]{pictures/1b_2.png} \includegraphics[width=50pt]{pictures/1b_3.png}\\ \textsc{t-} \textsc{o-} \textsc{m} \\ \\
\ex \label{ex:6d} Noun modified by index (\textsc{IX})\\\glll 
\includegraphics[width=50pt]{pictures/1d_1.png} \includegraphics[width=50pt]{pictures/1d_2.png}\\ \textsc{ix} \textsc{cat} \\ \\
\end{xlist}
\end{exe}
\end{multicols}
\ea 
\ea \label{ex:7} Pronominal \\\glll
\includegraphics[width=50pt]{pictures/2_1.png} \includegraphics[width=50pt]{pictures/2_2.png} \\ \textsc{ix} \textsc{sick} \\ \\
\z \z
\clearpage
\begin{multicols}{4}
\begin{exe}
\ex \label{ex:8}
\begin{xlist}
\ex Whole entity \\\glll 
\includegraphics[width=50pt]{pictures/3e.png} \\ \textsc{fish.fall}\\ \\
\ex Body part \\\glll
\includegraphics[width=50pt]{pictures/3b.png} \\ \textsc{walk} \\ \\ 
\ex Handling \\\glll 
\includegraphics[width=50pt]{pictures/3c.png} \\ \textsc{hold} \\ \\
\ex Extension \\\glll 
\includegraphics[width=50pt]{pictures/3d_1.png} \includegraphics[width=50pt]{pictures/3a.png}\\ \textsc{plank} {\textsc{dh: mouse}} \\ {} {\textsc{ndh: tunnel}}\\ \\
\end{xlist}
\end{exe}
\end{multicols}
\begin{multicols}{3}
\begin{exe}
\ex
\begin{xlist}
\ex \label{ex:9a} Constructed action \\\glll 
\includegraphics[width=50pt]{pictures/4c.png} \\ \textsc{struggle} \\ \\

\hypertarget{signed-vs.-spoken-res}{%
\subsection{Signed vs.~Spoken REs}\label{signed-vs.-spoken-res}}

Commonly in both spoken and sign languages nouns and noun phrases are
used as REs. Spoken languages use pronouns, determines and
demonstratives where sign languages use a pointing sign, usually glossed
as IX, which assumes similar functions (Lillo-Martin and Klima, 1990).
It is important to mention here that in some spoken languages (e.g.,
German and Dutch), demonstratives can also refer to persons (Bosch and
Umbach, 2007) and behave similarly to personal pronouns. In addition,
the pointing sign is used to referentially anchor a referent to a
location in signing space. For instance, when the signer introduces
\textsc{elvan} for the first time, s/he can assign her to the right.
This referential locus can later be used by agreement verbs and
classifier predicates (more below). For instance, in (\ref{ex:2a}) there
are two referents, \textsc{elvan}, who is referentially-assigned to the
right side of the signer, represented with the ``A'' subscript and
\textsc{demet}, who is referentially-assigned to the left side of the
signer, represented with the ``B'' subscript. When the signer makes a
contrast in (\ref{ex:2b}), as a pronominal use, she points to the right
side to refer to ELVAN and to the left to refer to \textsc{demet}. We
schematically represent the referential locus assigned to the referents
in (\ref{ex:3}).

@@

Grammatical gender as in Romance languages or noun classes as in Bantu
languages are some reference tracking tools that are familiar from
spoken languages. A somehow less known reference-tracking tool in spoken
languages is \emph{verbal classifiers}, which morphologically attach to
the verb and often cross-reference a core argument in a clause
(Aikhenvald, 2017). As outlined in the previous section, classifiers
denote sets of objects based on semantic criteria such as size, shape,
body-part, texture and sometimes function. Usually, the verbal
classifier and the cross-referenced noun co-occur in the same clause
when the argument is introduced into discourse. The classifier can track
the argument on its own on subsequent occurrences. In sign languages, a
verbal classifier morpheme is encoded by a \emph{handshape} that is
articulated phonologically co-temporally with a movement, which encodes
the verbal meaning (Benedicto and Brentari, 2004; Supalla, 1990). These
two infused morphemes form a complex predicate. Some classifier
handshapes and the set of objects that they denote are presented below
in (\ref{ex:4}).

@@

@@

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\vadjust pre{\hypertarget{ref-ahlgren1994}{}}%
Ahlgren, Inger, Bergman, Brita, 1994. Reference in narratives. I.
Ahlgren, B. Bergman \& M.

\leavevmode\vadjust pre{\hypertarget{ref-ahn2019}{}}%
Ahn, Dorothy, 2019. The determinacy scale: A competition mechanism for
anaphoric expressions (PhD thesis).

@@

\leavevmode\vadjust pre{\hypertarget{ref-hodge2019}{}}%
Hodge, Gabrielle, Ferrara, Lindsay N., Anible, Benjamin D., 2019. The
semiotic diversity of doing reference in a deaf signed language. Journal
of Pragmatics 143, 3353.

\leavevmode\vadjust pre{\hypertarget{ref-ilkbasaran2015}{}}%
Ilkbasaran, Deniz, 2015.
\href{https://www.proquest.com/docview/1656169535/abstract/16EA71911A644FC3PQ/1}{Literacies,
mobilities and agencies of deaf youth in turkey: Constraints and
opportunities in the 21st century} (PhD thesis). Ann Arbor, United
States.

@@

\leavevmode\vadjust pre{\hypertarget{ref-kemaloglu2012}{}}%
Kemaloğlu, Yusuf Kemal, Kemaloğlu, Pınar Yaprak, 2012. The history of
sign language and deaf education in Turkey. Kulak burun bogaz ihtisas
dergisi: KBB = Journal of ear, nose, and throat 22, 65--76.
doi:\href{https://doi.org/10.5606/kbbihtisas.2012.013}{10.5606/kbbihtisas.2012.013}

@@

\leavevmode\vadjust pre{\hypertarget{ref-ministryofnationaleducation2020}{}}%
Ministry of National Education, 2020.
\href{http://sgb.meb.gov.tr/meb_iys_dosyalar/2020_09/04144812_meb_istatistikleri_orgun_\%20egitim_2019_2020.pdf}{Milli
e{ğ}itim istatistikleri, örgün e{ğ}itim 2019/{'}20 {[}national education
statistics, formal education 2019/{'}20{]}}.
